Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Disparate Impact Testing}

### 2019-12-05 article-information-2019.tex
more progress on the text

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This text outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on realistic, publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and \textcolor{red}{global and local gradient-based or Shapley feature importance}. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using straightforward group fairness measures. By combining innovations in interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Even if responsible AI feels like a somewhat broad and amorphous notion, at least one aspect should be crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. Though this text makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), this text does propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems.

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this text with a brief outline of proposed additional steps to increase human trust and understanding in ML.

@@

To provide a sense of accuracy differences, this text compares the performance of more interpretable constrained ML models and less interpretable unconstrained ML models on simulated data and on collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a more realistic template workflow for future users of similar methods and tools.

@@

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in this text. Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing the same model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific and personalized adverse action notices for automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.

@@

MGBM and XNN interpretable model architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. These approaches are promising responses to the black-box problem in ML. However, they are just a small part of a burgeoning ecosystem of Python tools for interpretable models, post-hoc explanation, and discrimination testing and remediation. Any discussion of interpretable ML models would be incomplete without references to the seminal work of the Rudin group at Duke University and EBM or GA\textsuperscript{2}M models, pioneered by researchers Microsoft and Cornell and led by Rich Caruana. In keeping with a major theme of this text, models from these leading researchers and several other kinds of interpretable ML models are now available as open source Python packages. Among others, practitioners can now evaluate EBM in the \href{https://github.com/interpretml/interpret}{\texttt{interpret}} package, optimal sparse decision trees, GAMs in the \href{https://github.com/dswah/pyGAM}{\texttt{pyGAM}} package, a variant of Friedman's RuleFit  in the \href{https://github.com/scikit-learn-contrib/skope-rules}{\texttt{skope-rules}} package, monotonic calibrated interpolated lookup tables in \href{https://github.com/tensorflow/lattice}{\texttt{tensorflow/lattice}}, and ``this looks like that'' interpretable deep learning \cite{osdt}, \cite{rulefit}, \cite{lattice}, \cite{this_looks_like_that}.\footnote{Optimal sparse decision trees: \url{https://github.com/xiyanghu/OSDT}.}\textsuperscript{,}\footnote{``This looks like that'' interpretable deep learing: \url{https://github.com/cfchen-duke/ProtoPNet}.} Additional, relevant Python packages include: \href{https://github.com/dssg/aequitas}{\texttt{aequitas}} and \href{https://github.com/LASER-UMASS/Themis}{\texttt{Themis}} for discrimination testing, \href{http://aif360.mybluemix.net/}{\texttt{AIF360}} for discrimination testing and remediation, and \href{https://github.com/SeldonIO/alibi}{\texttt{alibi}} and \href{https://github.com/SauceCat/PDPbox}{\texttt{PDPbox}} for post-hoc explanation.\footnote{See: \texttt{\url{https://github.com/jphall663/awesome-machine-learning-interpretability}} for a longer, curated list of related software packages and resources.}

@@

%appeal: interpretable models + explanations

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:con}
% Model debugging
% Model management and monitoring
% Appeal
% local bias

@@

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/blueprint.png}
\label{fig:blueprint}
\caption{A diagram of a proposed holistic ML workflow in which explanations (highlighted in red) are used along with interpretable models, DI analysis and remediation techniques, and other review and appeal mechanisms to create an understandable and trustworthy ML system.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on realistic, publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and \textcolor{red}{global and local gradient-based or Shapley feature importance}. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using straightforward group fairness measures. By combining innovations in interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Even if responsible AI feels like a somewhat broad and amorphous notion, at least one aspect should be crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. Though this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), this text does propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems.

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this manuscript with a brief outline of proposed additional steps to increase human trust and understanding in ML.

@@

To provide a sense of accuracy differences, this manuscript compares the performance of more interpretable constrained ML models and less interpretable unconstrained ML models on simulated data and on collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a more realistic template workflow for future users of similar methods and tools.

@@

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in this manuscript. Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing the same model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific and personalized adverse action notices for automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.

@@

MGBM and XNN interpretable model architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. The same can be said of Shapley values and \textcolor{red}{discrimination tests}. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. While these approaches are promising responses to the black-box and discrimination problems in ML, they are just a small part of a burgeoning ecosystem of research and Python tools for responsible ML. Figure \ref{fig:blueprint} is a workflow blueprint that illustrates some of the additional steps that may be required to build a fully understandable and trustworthy machine learning system.\footnote{See: \texttt{\url{https://github.com/jphall663/hc_ml}} for details regarding the workflow in Figure \ref{fig:blueprint}.} While all the methods mentioned in Figure ~\ref{fig:blueprint} play an important role in increasing human trust and understanding of ML, a few pertinent references and Python resources are highlighted here as further reading.

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/blueprint.png}
\caption{A diagram of a proposed holistic ML workflow in which explanations (highlighted in red) are used along with interpretable models, DI analysis and remediation techniques, and other review and appeal mechanisms to create an understandable and trustworthy ML system.}
\label{fig:blueprint}
\end{figure}

Any discussion of interpretable ML models would be incomplete without references to the seminal work of the Rudin group at Duke University and EBM or GA\textsuperscript{2}M models, pioneered by researchers Microsoft and Cornell and led by Rich Caruana. In keeping with a major theme of this manuscript, models from these leading researchers and several other kinds of interpretable ML models are now available as open source Python packages. Among others, practitioners can now evaluate EBM in the \href{https://github.com/interpretml/interpret}{\texttt{interpret}} package, optimal sparse decision trees, GAMs in the \href{https://github.com/dswah/pyGAM}{\texttt{pyGAM}} package, a variant of Friedman's RuleFit  in the \href{https://github.com/scikit-learn-contrib/skope-rules}{\texttt{skope-rules}} package, monotonic calibrated interpolated lookup tables in \href{https://github.com/tensorflow/lattice}{\texttt{tensorflow/lattice}}, and ``this looks like that'' interpretable deep learning \cite{osdt}, \cite{rulefit}, \cite{lattice}, \cite{this_looks_like_that}.\footnote{Optimal sparse decision trees: \url{https://github.com/xiyanghu/OSDT}.}\textsuperscript{,}\footnote{``This looks like that'' interpretable deep learning: \url{https://github.com/cfchen-duke/ProtoPNet}.} Additional, relevant references and Python functionality includes:

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item \textbf{Exploratory data visualization (EDA)}: \texttt{H2OAggregatorEstimator} in \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2oaggregatorestimator}{\texttt{h2o}} \cite{wilkinson2018visualizing}.
\item \textbf{Sparse feature extraction}: \texttt{H2OGeneralizedLowRankEstimator} in \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogeneralizedlowrankestimator}{\texttt{h2o}} \cite{udell2016generalized}.
\item \textbf{Privacy preprocessing and private models}: differential privacy for ML in \href{https://github.com/IBM/differential-privacy-library}{\texttt{diffprivlib}} \cite{diffprivlib}, \cite{ji2014differential}; private models in \href{https://github.com/tensorflow/privacy/}{\texttt{tensorflow/privacy}} \cite{pate}, \cite{abadi2016deep}.
\item \textbf{Post-hoc explanation}: %\href{https://github.com/SeldonIO/alibi}{\texttt{alibi}} \href{https://github.com/SauceCat/PDPbox}{\texttt{PDPbox}} gradients
\item \textbf{Discrimination testing}: in \href{https://github.com/dssg/aequitas}{\texttt{aequitas}} and \href{https://github.com/LASER-UMASS/Themis}{\texttt{Themis}}.
\item \textbf{Discrimination remediation}: Reweighing, adversarial de-biasing, learning fair representations, and reject option classification in \href{http://aif360.mybluemix.net/}{\texttt{AIF360}} \cite{kamiran2012data}, \cite{zhang2018mitigating}, \cite{lfr}, \cite{kamiran2012decision}.
\item \textbf{Model debugging}: %tensorflow stuff cleverhans, foolbox, SALib what-if
\item \textbf{Model documentation}: Models cards \cite{model_cards}, e.g. \texttt{\url{https://github.com/openai/gpt-2}}, \texttt{\url{https://modelcards.withgoogle.com/object-detection}}. 
\end{itemize}

\noindent See: \texttt{\url{https://github.com/jphall663/awesome-machine-learning-interpretability}} for a longer, curated list of related software packages and resources.

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:con}

This text puts forward results on simulated data to provide a rough validation of constrained ML models, post-hoc explanation techniques, and discrimination testing methods. These same modeling, explanation, and discrimination testing approaches are then applied to more realistic mortgage data to provide an example of a responsible machine learning workflow. The discussed methodologies are solid steps toward interpretability, explanation, and minimal discrimination for ML decisions, which should ultimately enable increased fairness and logical appeal processes for ML decision subjects. Of course there is more to the responsible practice of ML than interpretable models, post-hoc explanation, and discrimination testing, even from a technology perspective, and this manuscript also points out numerous additional references and open source Python software assets that are available to researchers and practitioners today to increase human trust and understanding in ML systems. While the messy, complex, and human problems of racism, sexism, privacy violations, and cybercrime can probably not be solved by technology alone, this work and many others illustrate numerous ways for ML practitioners to become part of the solution to these problems, instead of perpetuating and exacerbating them.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer and investor of H2O.ai and a customer of BLDS, LLC.}
